Does Anybody Really Know What Time It Is?\\ Automating the Extraction of Date Scalars

### 2016-05-03 cikm2016/mdl.tex
More MDL verbiage

\subsection{Domains}
Potter's Wheel presents an algorithm for deriving a common structure for a set of strings by breaking each string down into a sequence of \textit{domains}. A domain is a set of strings, with a few optional statistical properties. They are defined by an interface that includes:
\begin{itemize}
\setlength\itemsep{0em}
\item A required inclusion function \functionname{match} to test for membership in the domain
(\eg the domain \functionname{<Digits>} returns true for \functionname{"123"} 
and false for \functionname{"abc"}.)
\item An optional function \functionname{cardinality} to compute the number of values in the domain with a given length
(\eg the domain for three letter month names (\functionname{MMM}) would return $12$ for the length $3$ and $0$ otherwise.)
\item An optional function \functionname{updateStatistics} to update statistics for the domain based on a given value
(\eg the domain \functionname{MMM} might keep a histogram of how frequently each month was encountered.)
\item An optional function \functionname{isRedundantAfter} to prevent consideration of a domain that is redundant.
(\eg  the domain \functionname{<Digits>} would be redundant after itself because there is no difference between a single run of digits and two adjacent runs.)
\end{itemize}

In our approach, we implement all of these functions, but with significant changes to the last one (described below) to increase the level of pruning applied during
domain enumeration.

@@

\subsection{Performance}
Structure enumeration is computationally expensive, so we added a number of enhancements to the original domain extraction algorithm to keep the run time low enough for interactivity.

@@

Date domains are also often \textit{uniform} in that adding more characters to a mismatch will not help. \eg a $2$-digit day domain that does not match a $1$-letter substring will not be able to generate a match by adding more characters.

\subsubsection{Parallel Evaluation}
We have also identified two opportunities for parallel computation during structure enumeration.
Enumerating the matching structures for a single sample is computationally expensive, but the samples themselves are independent. 
Partitioning the samples allows each thread to produce a set of candidate structures, which can eventually be merged to produce a single candidate list.
Then once generated, the evaluation of each independent structure (computation of the MDL, recording of domain statistics and parameterizing the structure) can be conducted in parallel.

\subsection{Unparameterization}
Domain parameterization is an important part of generating compact representations via MDL, but it creates problems for date recognition. For example, if a set of dates contains a constant month string (\eg all values are in September) it is important to keep track of the month name domain. 
Consequently, when we parameterize a constant generic \functionname{<Word>} domain, we tag it with any date part domain that it matched. 
We then need to apply an additional pruning step to remove any structures that also found an equivalent domain (\eg two-digit month). 
These rules are equivalent to the context-based redundancy rules in Section~\ref{subsec:mdl:redundancy}, but have to be applied again after the parameterization of generics.

\subsection{Global Pruning}
The pruning rules used for the structure extraction reduce the search space dramatically, but they are also contextual and can only look backwards. The domains also contain a fair amount of ambiguity that requires the application of domain knowledge. We therefore found it necessary to add some post-generation global pruning rules:
\begin{itemize}
\setlength\itemsep{0em}
\item The set of date parts cannot contain place value gaps (\eg structures that have year and day without month are removed.)
\item Similarly, the set of time parts cannot contain place value gaps and must also be in place value order (times are never written in orders such as \texttt{mhs}.)
\item The existence of time parts cannot make dates incomplete (\eg patterns like year-day-hour are removed.)
\item Two digit years require special handling. In particular, they cannot appear adjacent to a two-digit field if the structure contains any punctuation. (This can come up in some small early 21st century year domains where a two-digit year can masquerade as almost any numeric field \eg $08$ for $2008$.)
\end{itemize}

\subsection{Locale Sensitivity}
Providing an acceptable international user experience requires correctly handling the locale of the column text. Accordingly, at the start of the structure extraction we use the column locale to create a set of domains containing locale-sensitive strings such as month names. We also use the locale to map these strings to upper- and lower-case in addition to the ICU mixed case strings. This enables us to accurately compute MDL statistics without having to map the candidate strings at runtime (which would be slow).

Knowing the locale of a string is not always helpful. In our data set, we found numerous cases where the locale was specified (\eg Sweden) but the data was actually in English. Accordingly, we test both locales and rank the combined results.

@@

\subsection{Ranking}
MDL structure analysis produces a ranked list of format candidates, but we have found that a number of other properties of the formats should be preferred over simple compactness:
\begin{itemize}
\setlength\itemsep{0em}
\item Because we have a set of samples, we can apply the candidate format to the strings to see how well it performs. Formats with fewer parse errors are preferred.
\item Date parts can be considered a place-value system, so we prefer ``more significant'' components (\eg month-day-year over hour-minute-second).
\item If two formats from different locales give the same results, prefer the original column locale. The sample set may have missed an example where this could be important. For example, the month name for "September" is the same in both German and English, but ``October" and ``Oktober" are different.
\item If the format has an ambiguous date order (\eg all days are less than 12), then prefer the default date order of the locale. Again, the sample set may have missed a counterexample, so this is the best option.
\item Once these semantic preferences have been considered, we then prefer the more compact (MDL) representation.
\end{itemize}

>>>>

\subsection{Domains}
Potter's Wheel presents an algorithm for deriving a common structure for a set of strings by breaking each string down into a sequence of \textit{domains}. A domain is a set of strings, with a few optional statistical properties. Each one is defined by an interface that includes:
\begin{itemize}
\setlength\itemsep{0em}
\item A required inclusion function \functionname{match} to test for membership in the domain
(\eg the domain \functionname{<Digits>} returns true for \functionname{"123"} 
and false for \functionname{"abc"}.)
\item An optional function \functionname{cardinality} to compute the number of values in the domain with a given length
(\eg the domain for three letter month names (\functionname{MMM}) would return $12$ for the length $3$ and $0$ otherwise.)
\item An optional function \functionname{updateStatistics} to update statistics for the domain based on a given value
(\eg the domain \functionname{MMM} might keep a histogram of how frequently each month was encountered.)
\item An optional function \functionname{isRedundantAfter} to prevent consideration of a domain that is redundant.
(\eg  the domain \functionname{<Digits>} would be redundant after itself because there is no difference between a single run of digits and two adjacent runs.)
\end{itemize}

In our approach, we implement all of these functions, but with significant changes to \functionname{isRedundantAfter} (described below) to increase the level of pruning applied during
domain enumeration.

@@

\subsection{Performance}
Structure enumeration is computationally expensive, so we have added a number of enhancements to the original domain extraction algorithm to keep the run time low enough for interactivity.

@@

Date domains are also often \emph{uniform} in that adding more characters to a mismatch will not help. \eg a $2$-digit day domain that does not match a $1$-letter substring will not be able to generate a match by adding more characters.

\subsubsection{Parallel Evaluation}
We have also identified two opportunities for parallel computation during structure enumeration.
Enumerating the matching structures for a single sample is computationally expensive, but the samples themselves are independent. 
Partitioning the samples allows each thread to produce a set of candidate structures, which can eventually be merged to produce a single candidate list.
Once the list has been generated, the evaluation of each independent structure (computation of the MDL, recording of domain statistics and parameterizing the structure) can be conducted in parallel.

\subsection{Unparameterization}
Domain parameterization is an important part of generating compact representations via MDL, but it creates problems for date recognition. 
If (say) a set of dates contains a constant month string (\eg all values are in September) it is important to keep track of the month name domain
so that the \dateparse\ function will parse the month.
When we parameterize a constant generic \functionname{<Word>} domain, we therefore tag it with any date part domain that it matched. 
We then need to apply an additional pruning step to remove any structures that also found an equivalent domain (\eg two-digit month). 
These rules are equivalent to the context-based redundancy rules in Section~\ref{subsec:mdl:redundancy}, but have to be applied again after the parameterization of generics.

\subsection{Global Pruning}
The pruning rules used for the structure extraction reduce the search space dramatically, but they are also contextual and can only look backwards. 
The domains also contain a fair amount of ambiguity that requires the application of problem space knowledge. 
This made it necessary to add some post-generation global pruning rules:
\begin{itemize}
\setlength\itemsep{0em}
\item The set of date parts cannot contain place value gaps (\eg structures that have year and day without month are removed.)
\item Similarly, the set of time parts cannot contain place value gaps and must also be in place value order (times are never written in orders such as \functionname{mhs}.)
\item The existence of time parts cannot make dates incomplete (\eg patterns like year-day-hour are removed.)
\item Two digit years require special handling. In particular, they cannot appear adjacent to a two-digit field if the structure contains any punctuation. (This can come up in some small early 21st century year domains where a two-digit year can masquerade as almost any numeric field \eg $08$ for $2008$.)
\end{itemize}

\subsection{Locale}
Providing an acceptable international user experience requires correct handling of the column locale. 
Accordingly, at the start of the structure extraction we use the locale to create a set of domains containing locale-sensitive strings such as month names. We also use the locale to map these strings to upper- and lower-case in addition to the ICU mixed case strings. 
This enables us to accurately compute MDL statistics without having to map the input strings at runtime (which would be slow).

Knowing the locale of a string is not always helpful. In our test data, we found numerous cases where the locale was specified (\eg Sweden) but the data was actually in English. Accordingly, we test both locales and rank the combined results.

@@

\subsection{Ranking}
MDL structure analysis produces a ranked list of format candidates, but we have found that a number of other properties of the formats should be preferred over simple compactness:
\begin{itemize}
\setlength\itemsep{0em}
\item Because we have a set of samples, we can apply each format to the samples to see how well it performs. Formats with fewer parse errors are preferred.
\item Date parts can be considered a place-value system, so we prefer ``more significant'' components (\eg month-day-year over hour-minute-second).
\item If two formats from different locales give the same results, prefer the original column locale. The sample set may have missed an example where this could be important. For example, the month name for "September" is the same in both English and German, but ``October" and ``Oktober" are different.
\item If the format has an ambiguous date order (\eg all days are less than 12), then prefer the default date order of the locale. Again, the sample set may have missed a counterexample, so this is the best option.
\item Once these semantic preferences have been considered, we then prefer the more compact (MDL) representation.
\end{itemize}
